Trees

### 2019-01-24 Trees_making.Rmd
after spellcheck and minor corretions

```{r setup, cache = F, echo=FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
```

@@

The general flow of tree-making work is shown in the diagram below.  
You will usually start with species list from the mata-analytic data set (use binomial latin names for the species list and in your database). First, check if there is any phylogenetic tree available that already contains all your species (e.g. a supertree). Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic supertree from Open Tree of Life database (https://opentree.wikispaces.com/). We will use this approach in our examples below. Most of the time *rotl* will work well, especially with the common species. There are other super-trees available, e.g. for birds - we will use one of thes as our example too. Overall, if you can find a tree that contains all your specis, you just trim (subset) it down to your species. If a few species is missing, it is sometimes possible to substitute them with closely related species. If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:  
 - If the species are very diverged (e.g. mixture of vertabrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload lst of the species names from a file. This allows to generate a scaffold for a tree.  
 - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
 - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

If the resulting tree contains polytomies, these can be resolved at random, or based on information from more detailed trees (e.g. published phylogeneies of specific lower taxa). You can also graft missing species on your tree. Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage. Again, your final tree, must contain exactly the same list of species names as your meta-analytic data set.

```{r prepare, message = FALSE, echo = TRUE, eval = FALSE, warning = FALSE}
install.packages("diagram")
install.packages("tidyverse")
install.packages("ape")
install.packages("curl")
install.packages("fulltext")
install.packages("treebase")
install.packages("devtools")
devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
install.packages("rotl")
```

@@

We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life supertree, resolve polytomies, trim and plot trees and deal with some of the most common issues. We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree ditors, etc.) is also possible. We hope you get the idea of the main principles of the process.

@@

**Important**: use Latin names, not common names.   
**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different timepoints.

@@

Note that two of the species names in our list have typos, and are labelled TRUE in a column "approximate_match". Still, they were matched to correct species names.  
If typos are detected, it is worth correcting the names in the original dataset and re-running tree search.

@@

```{r rotl 10species tree, warning = FALSE}
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name") #we can get a Warning meassage here about collapsing single nodes
```

@@

Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria. Given we dont have any other bacteria in the dataset, and the position on the tree is correct, thats fine, we just need to rename the tree tip back to "Escherichia coli".

@@

There are a few main existing and established supertrees for birds, so they can be used to create a subtree containing only the species from the list.  
Here we will use on of these supertrees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448.). The website http://birdtree.org/ also has an online tool for subsetting their supertrees generating distribution of trees rather than one tree. For simplicity and demonstration puproses we will use R to subset (prune) a single bird supertree down to a smaller list of species.

@@

Prune supertree tree to a list of taxa from our list.

@@

**Notes**: We were lucky that all our species had matches on the supertree. Quite often there are no matches for some of the species names. In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the supertree. Finally, a sister species can be used when a species is truly missing. Remember to match the names on the final tree back to the original species list from the dataset.

@@

We will now handle a larger set of more closely related taxa, from a real data set.  
The data set we use the dataset *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.

@@

If you display the names (taxa), note that some alternative names were found. We asssume these are correct ones for now.  
Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.

@@

Its not binbary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
We will resolve them at random.

@@

1. Resolving one or more multifurcations in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html

@@

3. Resolving manually by useing information from known more specialised published phylogeny without polytomies (see next exercise for how to try to find these).

@@

As an exercise, we will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 from R.
However, Treebase sometimes may not connect or run the search string correctly. You can try search Treebase directly at www.treebase.org, but sometimes its quite slow or not working.
So, just in case we also provide the full tree files downloaded from there, which can be used in this exercise. In R, we first try to find a tree via Treebase website search - go to https://treebase.org/treebase-web/search/studySearch.html and enter Acer in teh search box and select "Abstract" from the drop-down menu next to it, press "Search" button. There are 4 relevant studies which potentially contain relevant trees - we will use the S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom. Then, go to the "Trees"" tab where you will se 2 trees from this study. We will use Tr418 - you can download a Nexus file by clicking the icons on the right. Not: you can start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even search for the most obscure species in the species tab - sometimes these approaches work better.

Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 dataset.

@@

Run the treebase search - get amx 50 trees that contain exactly these four species names.

@@

There are multiple trees available from several studies. Note that the lost of found results includes the tree 418 from study 342 we already identified in our Treebase website search.  
So, we will request specifically tree Tr418 Study S342 - Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist. It may take a moment, depending on the size of the tree. You can try requesting different trees later.

```{r dwonload single tree from treebase}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
Acers_tree <- Acers_tree[[1]] #extract tree from the list
```

If downloading from R treebase did not work, upload tree saved in a file:
```{r dwonload single tree from file, eval = FALSE}
Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
```

@@

### DISCUSSION
W hope you learn a few tricks on how to make phylogenetic-like trees of species that can be used in meta-analytic models. There is much more to learn, especially dealing with missing or rare species names, grafting trees together, resolving polytomies, plotting cool trees etc. In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimats of the species relationships, and as long as you get the key features right, the tree will be ok, and so your analyses (in most cases).

>>>>

```{r setup, cache = F, echo=FALSE, results = FALSE}
knitr::opts_chunk$set(error = TRUE) #allow some execution errors for demonstration purposes
knitr::opts_chunk$set(eval = TRUE, echo = TRUE, warning = FALSE)
sessionInfo()
```

@@

The general flow of tree-making work is shown in the diagram below.  
You will usually start with species list from the meta-analytic data set (use binomial Latin names for the species list and in your database). First, check if there is any phylogenetic tree available that already contains all your species (e.g. a super-tree). Currently, the easiest way to build a custom phylogenetic tree usable for meta-analysis is by using an R package called *rotl*, which accesses a synthetic super-tree from Open Tree of Life database (https://opentree.wikispaces.com/). We will use this approach in our examples below. Most of the time *rotl* will work well, especially with the common species. There are other super-trees available, e.g. for birds - we will use one of these as our example too. Overall, if you can find a tree that contains all your species, you just trim (subset) it down to your species. If a few species is missing, it is sometimes possible to substitute them with closely related species. If no existing phylogeny is readily usable, there are a few non-exclusive options depending on the range of species in the data set:  
 - If the species are very diverged (e.g. mixture of vertebrates and invertebrates), use NCBI Taxonomy Common Tree tool (http://www.ncbi.nlm.nih.gov/Taxonomy/CommonTree/wwwcmt.cgi). At the webpage, use an input box at the top of the page to search each species name or upload list of the species names from a file. This allows to generate a scaffold for a tree.  
 - If few major clades are present, assemble together sub-trees created from bigger trees, if these exist for these clades.    
 - Use phyloGenerator (http://willpearse.github.io/phyloGenerator/) or other phylogenetic software to create phylogeny based on DNA sequences, if a shared sequence set exists for a given set of species (check if the resulting phylogeny makes sense).

If the resulting tree contains polytomies, these can be resolved at random, or based on information from more detailed trees (e.g. published phylogenies of specific lower taxa). You can also graft missing species on your tree. Be aware of species names problems: typos, scientific name synonyms, subspecies - have to be re-checked / corrected / substituted if generate errors at any stage. Again, your final tree, must contain exactly the same list of species names as your meta-analytic data set.

```{r prepare, message = FALSE, echo = TRUE, eval = FALSE, warning = FALSE}
install.packages("metafor")
install.packages("diagram")
install.packages("tidyverse")
install.packages("ape")
install.packages("curl")
install.packages("fulltext")
install.packages("treebase")
install.packages("devtools")
devtools::install_github("ropensci/rotl", dependencies = TRUE, build_vignette=TRUE)
install.packages("rotl")
```

@@

We will give you some examples on how to use *rotl* package to download sub-trees from the Open Tree of Life super-tree, resolve polytomies, trim and plot trees and deal with some of the most common issues. We will try to do most of theses tasks in R, but doing many of these outside R (using websites, text and tree editors, etc.) is also possible. We hope you get the idea of the main principles of the process.

@@

**Important**: use Latin names, not common names.   
**Notes**: *rotl* crashes on some species lists. Also, Open Tree of Life database gets constantly updated with new info, so you can get different outputs at different time-points.

@@

Note that two of the species names in our list have typos, and are labelled TRUE in a column "approximate_match". Still, they were matched to correct species names.  
If typos are detected, it is worth correcting the names in the original data set and re-running tree search.

@@

```{r rotl 10species tree, warning = FALSE, results=FALSE}
tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]], label_format = "name") #we can get a Warning meassage here about collapsing single nodes
```

@@

Note that the species names have the Open Tree identifiers appended to them and *E.coli* was replaced with "mrcaott616ott617", which is non-taxon node within Bacteria. Given we dont have any other bacteria in the data set, and the position on the tree is correct, that's fine, we just need to rename the tree tip back to "Escherichia coli".

@@

There are a few main existing and established super-trees for birds, so they can be used to create a sub-tree containing only the species from the list.  
Here we will use on of these super-trees called "Ericson backbone", which is downloadable from http://birdtree.org/ (Jetz, W., G. H. Thomas, J. B. Joy, K. Hartmann, and A. O. Mooers. 2012. The global diversity of birds in space and time. Nature 491:444-448.). The website http://birdtree.org/ also has an online tool for subsetting their super-trees generating distribution of trees rather than one tree. For simplicity and demonstration purposes we will use R to subset (prune) a single bird super-tree down to a smaller list of species.

@@

Prune super-tree tree to a list of taxa from our list.

@@

**Notes**: We were lucky that all our species had matches on the super-tree. Quite often there are no matches for some of the species names. In such cases, first check the names for typos or unnecessary characters, then try to find synonymous names and check if these are included in the super-tree. Finally, a sister species can be used when a species is truly missing. Remember to match the names on the final tree back to the original species list from the data set.

@@

We will now handle a larger set of more closely related taxa, from a real data set.  
The data set we use the data set *dat.curtis1998* (contains studies on the Effects of Elevated CO2 Levels on Woody Plant Mass) included in the *metafor* package.

@@

If you display the names (taxa), note that some alternative names were found. We assume these are correct ones for now.  
Now that the taxon names are matched to the Open Tree identifiers, we can pass them to the function tol_induced_subtree to retrieve the relationships among these taxa.

@@

Its not binary -  polytomies are present within genera: Acer, Betula, Quercus and Fraxinus.
We will resolve them at random.

@@

1. Resolving one or more multifurcations (i.e. polytomies) in all possible ways: http://blog.phytools.org/2016/08/resolving-one-or-more-multifurcations.html

@@

3. Resolving manually by using information from known more specialized published phylogeny without polytomies (see next exercise for how to try to find these).

@@

As an exercise, we will try to find more resolved phylogeny for the four species from genus Acer that were included in Curtis1998 from R. 
However, Treebase sometimes may not connect or run the search string correctly. You can try search Treebase directly at www.treebase.org, but sometimes its quite slow or not working.
So, just in case we also provide the full tree files downloaded from there, which can be used in this exercise. To see more easily what is in the Treebase database, we first try to find a tree via Treebase website search - you can go to https://treebase.org/treebase-web/search/studySearch.html and enter "Acer" in the search box, then select "Title" from the drop-down menu next to it, press "Search" button.

There are 57 studies, but most of them were selected because they contain word "spaACER" (we are a bit unlucky with the taxon name). There are a few study titles that mention Acer which potentially may contain relevant trees and species: S2155, S2157 and S342. You can tick boxes next to them and then press button "Discard unchecked items" at the bottom of the window. Now, if you select just one of the remaining 3 studies and then go to the Taxa tab, you will see the list of taxa associated with a given study.

In this exercise we will use study S342 Ackerly D., Donoghue M.J.	1998, so check the box next to it and then "Discard unchecked items" button at the bottom. Then, go to the "Trees"" tab where you will see 2 trees from this study. We pick Tr418 (consensus, i.e. made from collapsing multiple possible similar single trees). You can download it as a Nexus file by clicking the icons on the right. Note: you can also start the search from Tree tab, by searching tree titles and then look at the studies from which these trees are, or even start the search for the most obscure species in the Species tab - sometimes these approaches may work better.

Now we will try to do similar work from R. First, we create a custom search string with our four Acer species names from Curtis1998 data set.

@@

Run the treebase search - get max 50 trees that contain exactly these four species names.

@@

There are multiple trees available from several studies. Note that the lost of found results includes the tree 418 from study 342 we already identified in our Treebase website search.  
So, we will request specifically tree Tr418 from Study S342 (the one we already found via Treebase website search; Ackerly D., Donoghue M.J.	1998	Leaf size, sapling allometry, and Corner's rules: phylogeny and correlated evolution in maples (Acer).	American Naturalist). It may take a moment, depending on the size of the tree. You can try requesting different trees later and comparing whether you get different sub-trees after pruning.

```{r dwonload one tree from treebase, results=FALSE}
Acers_tree <- search_treebase("418", by="id.tree", exact_match = TRUE) #returns a single tree in a list
Acers_tree <- Acers_tree[[1]] #extract tree from the list
```

If downloading from R treebase did not work, upload tree saved in a file:
```{r dwonload one tree from file, eval = FALSE}
Acers_tree <- read.nexus(file = "T418.nex") #read Nexus tree file
```

@@

### DISCUSSION
W hope you learn a few tricks on how to make phylogenetic-like trees of species that can be used in meta-analytic models. There is much more to learn, especially dealing with missing or rare species names, grafting trees together, resolving polytomies, plotting cool trees etc. In some cases getting "perfect" tree may not be possible - but phylogenies are always just estimates of the species relationships, and as long as you get the key features right, the tree will be OK, and so your analyses (in most cases).
